\documentclass[a4paper, 10pt]{article}
\usepackage[left=2.25cm,right=2.25cm,top=2.25cm,bottom=2.25cm]{geometry}
\usepackage{mathpazo}
\usepackage{amsfonts, amssymb, amsmath}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{xfrac}
\usepackage{lscape}
\usepackage[dvipsnames]{xcolor}
\usepackage{setspace}
\usepackage{subfig}
\usepackage{booktabs}
\usepackage{blindtext}
\usepackage{epsfig}
\usepackage{hyperref}
\usepackage{booktabs,caption,fixltx2e}
\usepackage[flushleft]{threeparttable}
\usepackage{outlines}
\newcommand{\beqn}{\begin{eqnarray*}}
	\newcommand{\eeqn}{\end{eqnarray*}}
\newcommand{\bi}{\begin{itemize}}
	\newcommand{\ei}{\end{itemize}}
\newcommand{\bnot}{\hat{\beta}_0}
\newcommand{\bone}{\hat{\beta}_1}

\newcommand{\claire}[1]{{\color{red} [CD: {#1}]}}

\geometry{ left = 1in, right = 1in, top =1in, bottom =1in }
\pagestyle{fancy}
\setlength{\headsep}{.4 in}
\setlength{\parskip}{0.2 cm}
\setlength{\parindent}{1cm}
\renewcommand{\headrulewidth}{.25pt}
\lhead{MQE: Economic Inference from Data \\Fall 2021}
\chead{}
\lfoot{}
\cfoot{}
\rfoot{\footnotesize \thepage}
\newenvironment{myindentpar}[1]%
 {\begin{list}{}%
         {\setlength{\leftmargin}{#1}}%
         \item[]%
 }
 {\end{list}}

\begin{document}


\vspace{10mm}
\linespread{1.25}
\bigskip
Dear MQE class of 2021,\\ 


Over the past couple years, this country has been engaged in long overdue questioning and self-examination on how racism is deeply embedded in our lives and behaviors. Like many institutions throughout the country, academic institutions are also taking part in this reflection. 

In the Economics field, events over the past two years  have given rise to conversations and self-reflection. Economics struggles with a diversity problem. White men continue to dominate the field. Black and Brown economists, and to a lesser extent women, continue to be underrepresented. Those that do enter the field face the constant battles, some large, some small, that come hand in hand with being a minority in one's place of work. This is particularly concerning given that economics is social science. How are we to provide scientific insights into our social systems and policies if we have only experienced society from a position of privilege?

I love this field, and as your professor, I hope I can communicate and transmit my passion to you. I believe that the econometric tool set that you are going to learn in this course is very powerful. Thus far, these powerful tools have often been yielded by privileged people with little insight into the lived truths of minorities. This has shaped the types of questions economists ask, the way we think about questions and the assumptions we make when answering them. Importantly, I fear that biases in our discipline go on to contaminate the industries and institution to which we send our graduates. 

Thus, in my continued effort to improve myself and to push for improvement in this field, I offer an apology, some promises, and a few requests.  


I apologize for mistakes I know I am going to make. For years I have been training to think and analyze as an economist. This training has been incredibly valuable, but I also know that it may lead me to fail at questioning certain problematic assumption that have been accepted as normal in this field.  I need your fresh eyes to help me see these. If a statement, question, or assumption strikes you as odd and questionable or not reflective of your experience, let's discuss it and work towards change rather than accepting silently. Furthermore, if I ever say or do something that you feel is inappropriate, please let me know. Give me the opportunity to apologize and learn from your experience so I can avoid making the same mistakes in the future. Know that it is coming from ignorance, not intent, and though the duty to speak up places a burden on you, future students will benefit from it.

I promise to listen. I have time set aside every week for private office hours. You can sign up via my teaching website (\href{https://sites.google.com/view/claireduquennois/teaching}{https://sites.google.com/view/claireduquennois/teaching}).  Stop by so I can get to know you one-on-one and we discuss anything you would like- complaints, personal struggles, advising, careers, research...- other than course materials and problem sets which can be addressed during my course office hours. 

Finally, I request that you listen, and listen equally. Too often the voices of under-represented minorities and women get drowned out. This can happen because of publication bias that fails to value research on minorities or the research of minority faculty. It can happen when researchers don't cite the work of minorities.  It can happen because when you are the only person of color in an office, or class, you have no one to talk to about your experiences. It can happen when otherwise well meaning individuals interrupt certain groups more frequently, or repeat a contribution or idea originally voiced by someone without crediting them. If I do this accidentally, let me know. If someone points out that you engaged in any of these behaviors, exercise grace: listen, apologize, and learn from it.  If a classmate stated an important point that got ignored, amplify their voice by restating it while crediting them. These behaviors can be hard to change, but with awareness and a concerted effort we can try. 
\vspace{3mm}

\noindent Sincerely, 


\noindent Dr. Claire Duquennois\\

\vspace{3mm}
\noindent \textbf{Diversity in Economics Resources: }

\noindent \textbf{The Sadie Collective:}\\
  \textit{"Addressing the pipeline and pathway problem for Black women in economics and related fields"}\\
\href{https://www.sadiecollective.org/}{https://www.sadiecollective.org/}

\noindent \textbf{Economists for Equity at Berkeley:} (Resources compiled by an Economics PhD Student group)\\
\href{https://berkeleyeconomistsforequity.weebly.com/}{https://berkeleyeconomistsforequity.weebly.com/}

\noindent \textbf{The AEA's SP:} (American Economic Association's Summer Program)\\
\textit{prepares talented individuals from under-represented groups for doctoral programs in Economics and related disciplines.} \\
\href{https://www.aeaweb.org/about-aea/committees/aeasp}{https://www.aeaweb.org/about-aea/committees/aeasp}

\noindent \textbf{The AEA's CSWEP:} (AEA's Committee on the Status of Women in the Economics Profession) \\
\href{https://www.aeaweb.org/about-aea/committees/cswep}{https://www.aeaweb.org/about-aea/committees/cswep}

\noindent \textbf{The AEA's CSMGEP:} (AEA's Committee on the Status of Minority Groups in the Economics Profession) \\
\href{https://www.aeaweb.org/about-aea/committees/cswep}{https://www.aeaweb.org/about-aea/committees/cswep}

\noindent \textbf{The AEA's CSLGBTQ:} (AEA's Committee on the Status of LGBTQ+ Individuals in the Economics Profession) \\
\href{https://www.aeaweb.org/about-aea/committees/aealgbtq}{https://www.aeaweb.org/about-aea/committees/aealgbtq}








\end{document}

